\vspace{1.0cm} \LARGE \bf Reassortment between influenza B lineages and the emergence of a co-adapted PB1-PB2-HA gene complex

### 2014-07-15 fluB.tex
Address #36
Slim down the discussion way down. Remove excessive/irrelevant
speculation.

\subsection*{Evidence of a co-adapted PB1-PB2-HA gene complex}
In this paper we show that the PB1, PB2 and HA segments of influenza B viruses are the only ones that have continuously maintained separate Vic and Yam lineages, while other segments have fixed either Vic or Yam lineages (Figures \ref{genomeGrid}, \ref{lineageRatiosOverTime} and \ref{railroadPlot}).
Evidence suggests that this is a result of prolonged lack of reassortment between Vic and Yam lineages in PB1, PB2 and HA (Figure \ref{betweenDiversity}) which possess co-assorting sequences detectable as high linkage disequilibrium (Figure \ref{segmentLD}).
The vast majority of the sampled evolutionary history of each segment of influenza B viruses since the split of Vic and Yam lineages has been spent in association with either completely Victoria or completely Yamagata lineage derived PB1-PB2-HA complexes (Figure \ref{stateTime}), suggesting that having `pure' lineage PB1-PB2-HA complexes is important for whole-genome fitness.
We propose that this pattern of coassortment is due to the action of selection and not simply biased or rare reassortment.

When comparing the mean tip-tip TMRCA deviations we find evidence of isolation by temporal distance - when reassortments occur between PB1, PB2 and HA segments those events tend to involve branches that have similar TMRCAs.
Other segment pairs, with the exception of MP-NA, seem to show little signal of this and are able to reassort with branches that have been evolving under temporally distant genomic backgrounds.
In addition, most strains with mixed-lineage PB1-PB2-HA complexes occurred in the early years of the Vic--Yam split, when the two lineages were presumably much more similar at the nucleotide and amino acid levels.
The most recent influenza B viruses with mixed-lineage PB1-PB2-HA complexes are B/Waikato/6/2005-like and B/Malaysia/1829782/2007-like viruses which circulated for only 5 months or are known from a single isolate, respectively.

We should clarify that when we say that Vic and Yam PB1-PB2-HA complexes of the influenza B virus are co-adapted and have co-evolved we mean that these 3 segments do not appear to be fit when derived from both lineages and that the evolution to this state has been intimate, in the sense that there has not been a successful reassortant combining these two lineages across great temporal distances, unlike the rest of the genome.

\subsection*{Causes of co-dependence between PB1-PB2-HA}
Previous studies have investigated possible co-dependence patterns between segments of influenza B viruses, by focusing on segments which would be expected to be co-adapted, \textit{e.g.} PB1-PB2-PA and HA-NA \citep{mccullers2004}.
Though it would be easy to explain co-adaptation between these segments by referring to their functional roles \textit{e.g.} PB1-PB2-PA form the polymerase heterotrimer and HA-NA have antagonistic activities, our findings suggest a counter-intuitive relationship between PB1, PB2 and HA segments.

It is perhaps easiest to explain co-dependence between PB1 and PB2 segments based on their functions as part of the influenza RNA-dependent RNA polymerase (RdRp) heterotrimer.
Indeed, trees of PB1 and PB2 segments exhibit high similarity in tip--tip TMRCAs (Figure \ref{deltaTMRCA}), suggesting highly similar TMRCAs.
In addition, PB1--PB2 reassortants are the rarest and least persistent among mixed-lineage PB1-PB2-HA strains and have not been isolated in great numbers.

Explaining the co-dependence of PB1+2 and HA segments is more difficult.
Many studies have noted a possible link between the HA, NA and PB1 segments of influenza A viruses \citep{bergeron2010,fulvini2011}.
A previously used technique for producing vaccine seed strains involved selecting for HA-NA reassortants, which often yielded PB1-HA-NA reassortants as a side-effect \citep{bergeron2010,fulvini2011}.
Recent experiments have suggested that the presence or absence of a `foreign' PB1 segment can have dramatic effects on HA concentration on the surface of virions and total virion production \citep{cobbin2013}.
Additional evidence for a relationship between PB1 and HA segments in influenza A viruses is given by previous influenza pandemics, which were caused by avian-human influenza A virus reassortants.
It has been established that at least for the 1957 and the 1968 influenza pandemics, caused by A/H2N2 and A/H3N2 subtypes, respectively, the viruses responsible were reassortants possessing PB1 and HA segments derived from avian influenza A viruses \citep{kawaoka1989}.
However, there have also been reassortant influenza A viruses circulating for prolonged periods of time in humans that did have disparate PB1 and HA segments, \textit{e.g.} H1N2 outbreaks in 2001 \citep{gregory2002} and H1N1/09 in 2009 \citep{smith2009}.

Another possibility is the action of frequency dependent selection in preserving the diversity in one segment, whilst the other segments hitchhike along.
A good candidate for this would be HA, as it is now the sole bearer of substantial antigenic diversity within the influenza B population, as the Vic lineage NA segment went extinct in 2002 (Figures \ref{genomeGrid}, \ref{tmrcaOT}, \ref{lineageRatiosOverTime} and \ref{railroadPlot}).
Previous research has found that avian influenza A virus HA and NA segments, which are the primary vehicles of antigenicity, exhibit vast diversity when compared to `internal segments' (including PB1 and PB2 segments), which show much more recent TMRCAs and less variability at the amino acid level \citep{chen2006,obenauer2006}.
This is easiest to interpret as frequency-dependent selection acting to preserve antigenic diversity \citep{worobey2014}.
If this is the case in influenza B viruses, we expect frequency-dependent selection to act on HA and indirectly, through some unknown association with HA, on PB1 and PB2 segments.

We find this hitchhiking scenario unlikely: if PB1+2 segments were hitchhiking with the HA segment stochastically we would expect to see more PB1+2 versus HA reassortants and fixation of Vic or Yam lineage PB1+2 segments in the influenza B population.
Our primary dataset comprising 452 complete influenza B virus genomes has one instance of these kinds of reassortants characterised by B/Bangkok/163/1990-like genome constellations.
The larger secondary dataset with 1603 complete genomes has one additional example of this: isolates with a B/Waikato/6/2005-like PB1-PB2-HA constellation.
We note that in all cases the genome constellations of these reassortants were not fixed in the influenza B population and the more recent B/Waikato/6/2005-like reassortants persisted for a much shorter period of time (5 months versus 4 years), suggestive of declining PB1+2/HA reassortant fitness over time.

It is possible that Vic and Yam lineages of PB1, PB2 and HA segments have simply drifted away from each other, without any one segment being the driver of diversity preservation in PB1, PB2 and HA segments.
In this case PB1, PB2 and HA segments of Vic and Yam lineage accumulate substitutions that improve their ability to co-operate with segments of the same lineage and worse at interacting segments of a different lineage.
This process, termed mutation-driven co-evolution \citep{presgraves2010}, has been suggested to be the cause of hybrid dysfunction in \textit{Saccharomyces} hybrids \citep{lee2008}.
It is widely accepted that Victoria lineage HA had been restricted to eastern Asia between 1992 and 2000 \citep{nerome1998,shaw2002}, offering a potential explanation for why the budding Victoria lineage segments were not homogenized via reassortment with Yamagata lineage segment in the early years of the split between the two lineages. 
The restriction to eastern Asia would presumably also give sufficient time for PB1, PB2 and HA segments to co-evolve together.
Similarly to the previous scenario, however, mutation-driven co-evolution would require an association of some kind between PB1+2 and HA segments to explain the low observed frequency and poor fitness of PB1+2/HA reassortants.
In addition, we see no reason why these three particular segments, and not the rest of the genome, would become co-adapted to each other in geographic isolation.

\subsection*{Suggested experiments}
The association between Victoria and Yamagata lineage PB1-PB2-HA complexes should be relatively straightforward to test in the lab.
Using previously developed plasmid systems \citep{hoffmann2002} it would be possible to create artificial reassortants, combining Vic and Yam lineages of PB1, PB2 and HA segments into mixed-lineage PB1-PB2-HA complexes.
We predict that artificially produced viruses with mixed-lineage PB1-PB2-HA complexes will have reduced fitness when compared to viruses with pure-lineage, i.e.\ entirely Vic or entirely Yam, PB1-PB2-HA complexes.
In addition, we expect the relationship between Vic and Yam lineage PB1, PB2 and HA segments to be dependent on date of segment isolation, as viruses with mixed-lineage PB1-PB2-HA complexes isolated earlier should perform better than viruses with PB1-PB2-HA segments isolated more recently.

Given the short circulation times of recent PB1-PB2-HA reassortants B/Waikato/6/2005-like and B/Malaysia/1829782/2007-like viruses we expect that artificially produced PB1-PB2 reassortants would be much less fit than either PB1-HA or PB2-HA reassortants.
We thus expect the following hierarchy of reassortant fitnesses (in order of decreasing fitness): PB1-PB2-HA, PB1-PB2/HA and PB1-HA/PB2 or PB2-HA/PB1, though there is some evidence to suggest that PB1-HA/PB2 might be more fit than PB2-HA/PB1 (Figures \ref{NormSPR_PB1-PB2_difference} -- \ref{NormSPR_PB2-HA_difference}).
The history of reassortments in influenza B viruses (Figure \ref{railroadPlot}) suggests that there are lineage-specific effects too, given the almost universal introgression of Yamagata lineage segments into Victoria PB1-PB2-HA background.
However, our analyses do not indicate any obvious differences in synonymous, non-synonymous or nucleotide substitution rates between Vic and Yam PB1-PB2-HA segment complexes or segments associated with either of the two (Figure \ref{robustCounting}).

However, we also see that epistatic effects might interfere with fitness measurements, if for example non-PB1-PB2-HA segments are also temporally mismatched.
Ideally, the co-adaptation would be easier to understand by referring to the structures of PB1 and PB2 proteins, as the link between these would be intuitive.
We have identified amino acid sites which are linked between PB1, PB2 and HA proteins of Victoria and Yamagata lineages, but we find very few sites on PB1 and PB2 proteins (Figure \ref{ChiCore}) that fall within the regions that form contacts within the influenza B polymerase heterotrimer \citep{sugiyama2009}, suggesting more subtle roles for sites we have identified.

\subsection*{The future of influenza B viruses}
We suggest that the preservation of two PB1-PB2-HA complex lineages is similar to genomic speciation islands, where small numbers of genes resist being homogenized through gene flow \citep{turner2005}.
In this context, we see three potential paths of evolution for influenza B viruses.
More segments could be recruited into the two currently circulating co-adapted segment complexes (PB1, PB2 and HA segments being the genomic speciation islands), as part of a speciation process, until all circulating influenza B viruses possess genomes with segments firmly associated with either the Vic or Yam lineage PB1-PB2-HA complex which could be referred to as belonging to either `new Victoria' or `new Yamagata' lineages.
This is the speciation scenario.
Due to the rarity of inter-lineage reassortment events it is unclear whether this process is already under way with NA and MP, based on Figures \ref{genomeGrid}, \ref{betweenDiversity} and \ref{deltaTMRCA}, being the next segments to be recruited to PB1-PB2-HA complexes.
The two alternatives to the speciation hypothesis are the `\textit{status quo}' model, where the influenza B genome continues to be homogenized via gene flow with the exception of PB1, PB2 and HA segments and the extinction scenario, whereby one of the two PB1-PB2-HA complexes goes extinct, marking the return of single-strain dynamics in the influenza B virus population.

Sympatric speciation in other systems usually requires strong barriers to introgression, \textit{e.g.} infertility of F1 hybrids can lead to the evolution of prezygotic reproductive isolation otherwise known as reinforcement or the Wallace effect.
To what extent this would apply to influenza B viruses remains unknown.
If influenza B viruses are undergoing sympatric speciation, it is imperative to determine whether co-infection with Victoria and Yamagata lineages of PB1-PB2-HA segments occur at a sufficiently high frequency and result in considerable losses of fitness to drive the evolution of reassortment isolation mechanisms (\textit{e.g.} unique packaging signals) or whether co-infection is so rare that speciation occurs via mutation-driven co-evolution.
We think that co-dependence between PB1 and PB2 segments can be explained by the fact that they are functionally linked: together with the PA protein they form part of the influenza B virus RNA-dependent RNA polymerase heterotrimer.

It also remains unknown whether reassortment events of the past 20 years (Figure \ref{railroadPlot}) which frequently involved the NP, MP and NS segments from a Yamagata PB1-PB2-HA background reassorting into a Victoria PB1-PB2-HA background are indicative of rare events followed by selective sweeps or stochastic fixation.
It's not unfeasible for selective sweeps following reassortments to break down developing co-adaptation of segments, especially if they are not functionally linked and have themselves been reassorted into a new PB1-PB2-HA background recently.
This is the second scenario we might expect to occur, whereby relatively frequent reassortments occur between Vic and Yam lineages and are followed by selective sweeps.
In this case influenza B viruses would undergo periodic genome homogenization events with the exception of PB1, PB2 and HA segments.
Because Vic HA and Yam HA are antigenically dissimilar we think that frequency-dependent selection would prevent even strong selective sweeps from driving the opposing PB1-PB2-HA gene complex to extinction.
This `\textit{status quo}' model would require strong selective sweeps and/or relatively frequent reassortments, neither of which seem to be lacking in the influenza B virus population.

Given the relatively recent explosion of sequence data available for influenza B, it is difficult to say whether dynamics similar to Victoria and Yamagata lineages have not occurred in influenza B virus genomes before and left no trace through extinction.
We find it unlikely that either Victoria or Yamagata lineage PB1-PB2-HA complexes will go extinct stochastically in the near future, as they have co-circulated for prolonged periods at a ratio close to 0.5, suggesting the action of frequency dependent selection (Figure \ref{lineageRatiosOverTime}).
Extinction through depletion of susceptible individuals, such as following influenza pandemics or mass vaccination seem unlikely as well.
Both Victoria and Yamagata lineage PB1-PB2-HA complexes survived the admittedly mild influenza pandemic in 2009 and influenza vaccines, which are usually applied to specific subsets of the population, do not produce lifelong immunity.

\subsection*{Conclusion}
We have used the $\dtmrca$ statistic to determine the degree of similarity in TMRCA dates between two temporally calibrated phylogenies.
We believe that patristic distance methods such as this, though themselves far from being new, have considerable power to address a wide variety of problems when combined with temporal phylogenies.
One of many useful applications of the $\dtmrca$ method would be identifying clades or taxa that are products of reticulate evolution.
$\dtmrca$ measures between independent analyses of the same alignment (`within-alignment') could be used as a cutoff to detect outlier taxa with greater than expected `between-alignment' $\dtmrca$ values.
To develop further, however, the statistical properties of patristic distance methods have to be evaluated in greater detail.
In addition, by treating the relative position of each isolate within a phylogeny of one segment as a label and modelling it on phylogenies of other segments we have also developed a metric similar to `between population' diversity used in calculating $F_{ST}$, which is capable of quantifying reticulate evolution-induced loss of diversity between partitions of taxa in two or more phylogenies.

>>>>

\subsection*{The origins of the linkage remain unknown}
The origin of the strong genetic linkage between PB1, PB2 and HA segments remains unclear.
We believe there are two alternative, but similar explanations for the origins of the strong genetic linkage between PB1, PB2 and HA segments: mutation-driven co-evolution \citep{presgraves2010} and Dobzhansky-Muller incompatibility \citep{dobzhansky1937,muller1942}.
Mutation-driven co-evolution \citep{presgraves2010}, has been suggested to be the cause of hybrid dysfunction in \textit{Saccharomyces} hybrids \citep{lee2008}, and evolves as a by-product of adaptation.
If one or the other influenza B lineage has undergone adaptation we might expect these changes to be beneficial in the derived background and incompatible with the ancestral background.
Dobzhansky-Muller incompatibility operates in a similar way, but the main difference from the scenario described earlier is that the incompatible alleles are neutral or nearly neutral in their native background and become deleterious or lethal when combined with non-native backgrounds.

\subsection*{Potential mechanisms for reassortant incompatibility}
Unfortunately, the limited amount of genomic data available for the early years of the Vic--Yam split precludes any attempts at answering whether selection or drift have led to the current linkage of PB1, PB2 and HA segments.
Although the origins of the linkage between these three segments might be difficult to explain, we can speculate on the nature of reassortant incompatibility.
For example, it is intuitive for why this might be the case for PB1 and PB2: both proteins operate cooperatively in intimate proximity as part of the RNA-dependent RNA polymerase heterotrimer.
Indeed, we observe that PB1--PB2 reassortants are the rarest and least persistent among mixed-lineage PB1-PB2-HA strains and have not been isolated in great numbers.
In fact most reassortants breaking the PB1-PB2-HA complex apart have occured in the past, close to the split of Vic and Yam lineages and are very rare.

Although the linkage between the two polymerase segments and HA is more difficult to explain, it is widely accepted that Victoria lineage HA had been restricted to eastern Asia between 1992 and 2000 \citep{nerome1998,shaw2002}, offering ample time for the budding Victoria lineage to develop reassortant incompatibility alleles.
There is some evidence that the linkage between PB1 and HA might not be a phenomenon restricted to influenza B viruses.
A technique used previously for producing vaccine seed strains involved selecting for HA-NA reassortants, which often yielded PB1-HA-NA reassortants as a side-effect \citep{bergeron2010,fulvini2011}.
Recent experiments have suggested that the presence or absence of a `foreign' PB1 segment can have dramatic effects on HA concentration on the surface of virions and total virion production \citep{cobbin2013}.
Additional evidence for a relationship between PB1 and HA segments in influenza A viruses is given by previous influenza pandemics, which were caused by avian-human influenza A virus reassortants.
It has been established that at least for the 1957 and the 1968 influenza pandemics, caused by A/H2N2 and A/H3N2 subtypes, respectively, the viruses responsible were reassortants possessing PB1 and HA segments derived from avian influenza A viruses \citep{kawaoka1989}.
However, there have also been reassortant influenza A viruses circulating for prolonged periods of time in humans that did have disparate PB1 and HA segments, \textit{e.g.} H1N2 outbreaks in 2001 \citep{gregory2002} and H1N1/09 in 2009 \citep{smith2009}.

The association between Victoria and Yamagata lineage PB1-PB2-HA complexes should be relatively straightforward to explore in the lab.
Using previously developed plasmid systems \citep{hoffmann2002} it would be possible to create artificial reassortants, combining Vic and Yam lineages of PB1, PB2 and HA segments into mixed-lineage PB1-PB2-HA complexes.
There is some evidence, based on the frequency and persistence times of different reassortant classes we have observed, to suggest that there should be a hierarchy of reassortant fitness starting with PB1+PB2+HA reassortants which should be the most fit, followed by PB1+2/HA, then PB1+HA/PB2 and finally PB2+HA/PB1 reassortants with the lowest fitness.
We believe that this is the most direct approach to unravelling the mechanism responsible for the linkage within the PB1-PB2-HA complex.

\subsection*{Will influenza B viruses speciate?}
We suggest that the preservation of two PB1-PB2-HA complex lineages is similar to genomic speciation islands, where small numbers of genes resist being homogenized through gene flow \citep{turner2005}.
In this context, we see three potential paths of evolution for influenza B viruses.
If more segments get recruited to the PB1-PB2-HA complex, the process could continue until none of the segments are able to reassort across the Victoria--Yamagata lineage boundary.
The two alternatives to the speciation hypothesis are the `\textit{status quo}' model, where the influenza B genome continues to be homogenized via gene flow with the exception of PB1, PB2 and HA segments and the extinction scenario, whereby one of the two PB1-PB2-HA complexes goes extinct, marking the return of single-strain dynamics in the influenza B virus population.
The eventual fate of influenza B viruses will be mostly determined by the combined effects of reassortment frequency and epistatic interactions between segments over long periods of time.

\section*{Conclusion}
We have used the $\dtmrca$ statistic to determine the degree of similarity in TMRCA dates between two temporally calibrated phylogenies.
We believe that patristic distance methods such as this, though themselves far from being new, have considerable power to address a wide variety of problems when combined with temporal phylogenies.
One of many useful applications of the $\dtmrca$ method would be identifying clades or taxa that are products of reticulate evolution.
$\dtmrca$ measures between independent analyses of the same alignment (`within-alignment') could be used as a cutoff to detect outlier taxa with greater than expected `between-alignment' $\dtmrca$ values.
To develop further, however, the statistical properties of patristic distance methods have to be evaluated in greater detail.
In addition, by treating the relative position of each isolate within a phylogeny of one segment as a label and modelling it on phylogenies of other segments we have also developed a metric similar to `between population' diversity used in calculating $F_{ST}$, which is capable of quantifying reticulate evolution-induced loss of diversity between partitions of taxa in two or more phylogenies.
